Wide binaries in Gaia DR1

### 2016-09-25 paper/ms.tex
radial velocity

% define macros for text
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\figname}{Figure}

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transpose}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}

@@

>>>>

% define macros for text
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\rave}{\project{\acronym{RAVE}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\figname}{Figure}

% define macros for math
\newcommand{\given}{\,|\,}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\transp}[1]{{#1}^{\mathsf{T}}}
\newcommand{\inv}[1]{{#1}^{-1}}
\newcommand{\bs}[1]{\boldsymbol{#1}}
\newcommand{\vperp}{\bs{v}^\perp}
\newcommand{\propm}{\bs{\mu}}
\newcommand{\matrx}[1]{\mathbf{#1}}
\newcommand{\kms}{\rm km~s^{-1}}

@@

\section{Methods} \label{sec:methods}

For a pair of stars $(i,j)$, we want to ratio the fully marginalized likelihood
(FML) of the stars having the same 3-space velocity, $\bs{v}=(v^\perp_\alpha,
v^\perp_\delta, v_r)$, compared to the FML of the stars having different 3-space
velocities:
\begin{equation}
  \frac{p(\propm_i, \varpi_i, \matrx{C}_i, \propm_j, \varpi_j, \matrx{C}_j)_{\bs{v}_i = \bs{v}_j}}
  {p(\propm_i, \varpi_i, \matrx{C}_i, \propm_j, \varpi_j, \matrx{C}_j)_{\bs{v}_i \neq \bs{v}_j}}.
\end{equation}
When this ratio is large---that is, the likelihood that the stars have the same
3-space velocity is large and the the likelihood that the stars have different
3-space velicities is small---the pair is a candidate binary star. For \gaia\
DR1, we only observe a parallax, $\varpi$, and two proper motion components,
$\propm = (\mu_\alpha, \mu_\delta)$. Where there is overlap with \rave\ or
\apogee, some number of observed radial velocities, $\tilde{v}_r$, with observed
variances $\sigma^2_{v_r}$ (independent of the \gaia\ data) may also be
available. The likelihood expressions above are marginalized over true 3-space
velocity and distance for each star in the pair. For the numerator,
\begin{equation}
  p(\propm_i, \varpi_i, \matrx{C}_i, \propm_j, \varpi_j, \matrx{C}_j)_{\bs{v}_i = \bs{v}_j} =
    \int \, \dd \bs{v} \, \dd d_i \, \dd d_j \,
    p(\propm_i, \varpi_i, \bs{v}, d_i, \matrx{C}_i) \,
    p(\propm_j, \varpi_j, \bs{v}, d_j, \matrx{C}_j)
\end{equation}
where
\begin{align}
  p(\propm, \varpi, \bs{v}, d, \matrx{C}) &=
    p(\propm, \varpi \given \vperp, d, \matrx{C}) \, p(\vperp, v_r) \, p(\tilde{v}_r \given v_r, \sigma^2_{v_r}) \, p(d \given l,b) \\
  p(\propm, \varpi \given \vperp, d, \matrx{C}) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{\left(\varpi, \mu_\alpha, \mu_\delta\right)} \\
  \bs{x}_\theta &= \transp{\left(\frac{1}{d}, \frac{v^\perp_\alpha}{d}, \frac{v^\perp_\delta}{d}\right)}
\end{align}

We assume Gaussian priors on the true velocity components
\begin{align}
  p(\bs{v}) = \mathcal{N}(0,25)~\kms
\end{align}
and take one of the distance priors from Coryn Bailer-Jones,

For the denominator of the likelihood ratio,
\begin{multline}
  p(\propm_i, \varpi_i, \matrx{C}_i, \propm_j, \varpi_j, \matrx{C}_j)_{\bs{v}_i \neq \bs{v}_j} = \\
    \int \, \dd \bs{v}_i \, \dd \bs{v}_j \, \dd d_i \, \dd d_j \,
    p(\propm_i, \varpi_i, \bs{v}_i, d_i, \matrx{C}_i) \,
    p(\propm_j, \varpi_j, \bs{v}_j, d_j, \matrx{C}_j)
\end{multline}
